---
title: 'r2mlm: R-squared for Multilevel Models'
author: "Mairead Shaw"
output: html_document
---

## Load Data and Dependencies

```{r message=FALSE, warning=FALSE}
library(r2mlm) # this also loads lme4
library(lmerTest) # significance for coefficients
library(performance)
library(dplyr)
```

## Null Model

Teachers clustered within classes:

```{r}
null_model <- lmer(satisfaction ~ 1 + (1|schoolID), data = teachsat, REML = TRUE)
summary(null_model)

r2mlm(null_model)
```

Three sets of output: Decompositions, R2s, and Graph

* Decompositions give you the unique R-squareds
* R2s gives you the unique R-squareds and the combinations
* Graph visualizes Decompositions

My order of reading is usually: R2s, then look at the graph.

The only variance explained in job satisfaction is explained by group membership (random intercept). That's the intraclass correlation! So 31.6% of variance in job satisfaction is attributed to cluster. For comparison, we can use the performance package to calculate the ICC:

```{r}
performance::icc(null_model)
```

They match, as expected!

## Full Model

1. Level-1: salary (centered within school)
2. Level-2: student-teacher ratio (same for all teachers within a school, differs across schools)

```{r}
model <- lmer(satisfaction ~ 1 + salary_c + s_t_ratio + (salary_c|schoolID), 
              data = teachsat, 
              REML = TRUE,
              control = lmerControl(optimizer = "bobyqa")) # optimizer change to help convergence (to make the example cleaner)
summary(model) # Instructive to look at our unstandardized results, too: coefficient values

r2mlm(model)
```

## Manual Entry

```{r}
r2mlm_manual(
  data = teachsat,
  within_covs = c("salary_c"),
  between_covs = c("s_t_ratio"),
  random_covs = c("salary_c"),
  gamma_w = c(0.074648),
  gamma_b = c(7.189783, -0.037282),
  Tau = as.matrix(bdiag(VarCorr(model))),
  sigma2 = getME(model, "sigma")^2,
  has_intercept = TRUE,
  clustermeancentered = TRUE
)
```

Compare to the automatic entry: they're the same!

## Model Comparison

```{r}
r2mlm_comp(null_model, model)
```

5 graph outputs: model A and B each overall, then comparisons of within, between, and total.

There is also manual entry for model comparison, but it's verbose so we're not going to look at it here.

## Non-CWC Model Options

```{r}
# You can see the centering, each school has a mean of zero
teachsat %>% 
  group_by(schoolID) %>% 
  summarize(
    mean(salary_c)
  )

# Remove centering on salary by adding a constant to each value
teachsat <- teachsat %>%
  mutate(salary = salary_c + 1)

# You can see mean is no longer zero
teachsat %>% 
  group_by(schoolID) %>% 
  summarize(
    mean(salary)
  )

# Model with this new salary
model_uncwc <- lmer(satisfaction ~ 1 + salary + s_t_ratio + (salary|schoolID), 
              data = teachsat, 
              REML = TRUE,
              control = lmerControl(optimizer = "bobyqa")) 
summary(model_uncwc)
```

If you use the regular `r2mlm` function, you get just an overall breakdown. You can use `r2mlm_long` instead to get a full breakdown, though this option is only available as a manual function right now.

```{r}
r2mlm_long_manual(
  data = teachsat,
  covs = c("salary", "s_t_ratio"),
  random_covs = c("salary"),
  clusterID = "schoolID",
  gammas = c(0.074648, -0.037282),
  Tau = as.matrix(Matrix::bdiag(VarCorr(model_uncwc))),
  sigma = getME(model_uncwc, "sigma")^2,
  bargraph = TRUE
)
```

Three outputs: Decompositions, R2s, Graph. This is similar to r2mlm(), but now there are two v terms. Recall that v is variance attributed to random slopes: v1 is slope variation within a cluster; v2 is slope variation between clusters. Why two v's here? When a level-1 variable is centered within cluster, every cluster has a mean of 0 and there is no variance between clusters. All the variance is within clusters. When a level-1 variable is NOT centered within cluster, it varies both within and between clusters.

What does the `r2mlm_long` output look like with our centered model?

```{r}
summary(model)

r2mlm_long_manual(
  data = teachsat,
  covs = c("salary_c", "s_t_ratio"),
  random_covs = c("salary_c"),
  clusterID = "schoolID",
  gammas = c(0.074648, -0.037282),
  Tau = as.matrix(Matrix::bdiag(VarCorr(model))),
  sigma = getME(model, "sigma")^2,
  bargraph = TRUE
)
```

The same, as expected! The change: v2 is essentially zero, because there is no variance in salary_c across groups.

